---
title: "Life Expectancy..."
subtitle: "  "
author: 
  - Karmen Zhu, Edan Wong, Saanvi Prasanth
thanks: "Code and data are available at: _______"
date: 30 September 2024
date-format: long
abstract: "  "
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
data <- LifeExpectancyData
# Rename the column using colnames()
colnames(data)[colnames(data) == "Life expectancy"] <- "LifeExpectancy"
colnames(data)[colnames(data) == "Income composition of resources"] <- "IncomeComposition"
#data <- read.csv('~/STA302-group55/LifeExpectancy/LifeExpectancy/data/raw_data')
```


# Introduction



# Data {#sec-data}

## Raw Data {#sec-first-point}
The data used in this paper is access in from Open Data Toronto and the particular data set used was the Daily Shelter & Overnight Service Occupancy & Capacity (@opendatatoronto). To analysis the data and creating graphs using the data, following package that was build in the (R program @citeR) was used: tidyverse (@tidyverse), dplyr (@dplyr), lubridate (@lubridate), and ggplot2 (@ggplot2). We clean the column names, separate the date into `year` and `month`, and create a standardized date column for monthly aggregation. The cleaned data is then saved for further analysis.

```{r}
# Randomly select 10 rows from the data
sampled_data <- data |> sample_n(10)

# Display the selected data
sampled_data
```

## Limitation of Data



# Results {#sec-mydatasection}

## Response variable(Life Expectancy)
```{r}
#| label: fig-response-variable
#| fig-cap: false
#| echo: false

morandi_hist <- rgb(0.2, 0.0, 0.0, 0.4)
  hist(data$LifeExpectancy,
     main = "Distribution of Life Expectancy",
     xlab = "Life Expectancy",
     col = morandi_hist,
     breaks = 10)
```
(explain why it fits linear regression, normality)


```{r}
#| label: fig-responsepredictor-plots
#| fig-cap: Visualizing the Relationship Between Life Expectancy and Its Predictors
#| echo: false

par(mfrow = c(2, 3))

# 1. Life Expectancy VS GDP plot
morandi_color_gdp <- rgb(0.4, 0.6, 0.4, 0.4)
plot(data$GDP, data$LifeExpectancy, 
     main = "Life Expectancy VS GDP", 
     xlab = "GDP", 
     ylab = "Life Expectancy",
     col = morandi_color_gdp,
     pch = 19,
     cex = 0.5)
# Add the regression line in red
abline(149.7477, 0.2592, col = "red", lty = 2)

# 2. Life Expectancy VS Polio Immunization Coverage plot
morandi_color_polio <- rgb(0.8, 0.6, 0.6, 0.4)
plot(data$Polio, data$LifeExpectancy, 
     main = "VS Polio Immunization Coverage", 
     xlab = "Polio", 
     ylab = "Life Expectancy",
     col = morandi_color_polio,
     pch = 19,
     cex = 0.5)
# Add the regression line in red
abline(149.7477, 0.2592, col = "red", lwd = 2)

# 3. Life Expectancy VS Diphtheria Immunization Coverage plot
morandi_color_diphtheria <- rgb(0.6, 0.8, 0.8, 0.4)
plot(data$Diphtheria, data$LifeExpectancy, 
     main = "VS Diphtheria Immunization Coverage", 
     xlab = "Diphtheria", 
     ylab = "Life Expectancy",
     col = morandi_color_diphtheria,
     pch = 19,
     cex = 0.5)
# Add the regression line in red
abline(149.7477, 0.2592, col = "red", lwd = 2)

# 4. Linear Model for Life Expectancy by Developed/Developing Status
# Create dummy variables for species
StatusDeveloped = as.integer(data["Status"] == "Developed")
StatusDeveloping = as.integer(data["Status"] == "Developing")

df_subsetted = data[, c(3, 4)]
df_converted = data.frame(df_subsetted, StatusDeveloped, StatusDeveloping)
fit_manually = lm(LifeExpectancy ~ StatusDeveloped + StatusDeveloping, data = df_converted)
summary(fit_manually)

# 5. Life Expectancy VS Income Composition of Resources plot
morandi_color_income <- rgb(0.6, 0.6, 0.8, 0.4)
plot(data$IncomeComposition, data$LifeExpectancy, 
     main = "VS Income Composition of Resources", 
     xlab = "Income Composition of Resources", 
     ylab = "Life Expectancy",
     col = morandi_color_income,
     pch = 19,
     cex = 0.5)
# Add the regression line in red
abline(149.7477, 0.2592, col = "red", lty = 2)

# Reset layout back to normal (1 plot per page)
par(mfrow = c(1, 1))

```



```{r}
#| label: fig-TotalExpenditure
#| fig-cap: Preliminary Model using the Predictors
#| echo: false
fit = lm(LifeExpectancy ~  GDP + Polio + Diphtheria + Status + IncomeComposition, data=data)
fit
```
```{r}
#| label: fig-cor-matrix
#| fig-cap: Correlation Matrix
#| echo: false
# Convert Status to a numeric variable
data$Status <- as.numeric(factor(data$Status))

cor_data <- data[, c("LifeExpectancy", "GDP", "Polio", "Diphtheria", "Status", "IncomeComposition")]
# correlation matrix
cor_matrix <- cor(cor_data, use = "complete.obs")
print(cor_matrix)
```
```{r}
#| label: fig-cor-plot
#| fig-cap: Correlation Plot
#| echo: false

library(PerformanceAnalytics)
library(dplyr)
cor_data <- data[, c("LifeExpectancy", "GDP", "Polio", "Diphtheria", "IncomeComposition")]
chart.Correlation(cor_data, histogram = TRUE, pch = 19)
```

```{r}
#| label: fig-residual-fitted
#| fig-cap:false
#| echo: false
morandi_residual_fitted <- rgb(1.0, 0.8, 0.4, 0.4)
plot(fit$fitted.values, fit$residuals,
     main = "Residuals vs Fitted",
     xlab = "Fitted values",
     ylab = "Residuals",
     pch = 19, col = morandi_residual_fitted)
abline(h = 0, col = "red", lty = 2)
```


```{r}
#| label: fig-TotalExpenditure
#| fig-cap: Partial Residual Plots
#| echo: false

library(car)
fitted <- lm(LifeExpectancy ~ GDP + Polio + Diphtheria + IncomeComposition, data = data)
crPlots(fitted, smooth = list(span = 0.8))  # 'span' is for stability
```



```{r}
#| label: fig-qq
#| fig-cap: false
#| echo: false

morandi_qq <- rgb(0.4, 0.2, 0.2, 0.4)
qqnorm(fit$residuals, main = "Q-Q Plot of Residuals", col = morandi_qq)
qqline(fit$residuals, col = "red", lty = 2)
```
```{r}
#| label: fig-scale-location
#| fig-cap: Scale-Location: Standardized Residuals vs Fitted Values
#| echo: false

morandi_scale <- rgb(0.0, 0.4, 1.0, 0.4)
standardized_residuals <- rstandard(fit)
plot(fit$fitted.values, sqrt(abs(standardized_residuals)),
     main = "Scale-Location",
     xlab = "Fitted Values",
     ylab = "√|Standardized Residuals|",
     pch = 19, col = morandi_scale)

# Add a smoothing line to check for patterns
lines(lowess(fit$fitted.values, sqrt(abs(standardized_residuals))), col = 'red', lwd = 2)
```


```{r}
#| label: fig-leverage
#| fig-cap: Scale-Location: Standardized Residuals vs Fitted Values
#| echo: false

morandi_leverag <- rgb(0.6, 0.2, 0.0, 0.4)
plot(fit, which = 5, main = "Residuals vs Leverage Plot", col = morandi_leverag)
```

\newpage

# Discussion

## First discussion point
From the visualizations (@fig-beds-status), 

## Second discussion point
As shown in the graph (@fig-service-type), 

## Third discussion point
From(@fig-homeless-groups), 


## Weaknesses and next steps


# Conclusion



\newpage

\appendix

# Appendix {-}


# Additional data details



\newpage


# References


